Drought-sensitive trees survive on deep water access in a tropical forest community

### 2020-08-03 manuscript.rmd
minor edits

\
**Corresponding author**: 
Rutuja Chitra-Tarak, A mispelt word.
\
E: arutuj@gmail.com
\
T: +1(505) 665 1647

@@

\
Trade-offs shape how trees' make a living in their environment (ecological strategies) and also provide Dynamic Vegetation Models (DVMs) the essential constraints to represent species-rich communities under changing environment [@Koven:2020]. This entails indentifying trade-offs among environment, traits and demography. A universal 'fast-slow' (aka productive-persistent, or acquisitive-conservative) plant economics spectrum conceptualises plants coordination for all resources and organs, both above- and below-ground [@Reich:2014]. For water resource observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified a fast-slow spectrum, although examples of all trade-offs observed simultaneously are rare, and belowground trait or environment linkages are weak. 'Fast' species at the productive end of the spectrum have traits with high capacity to acquire, move, store and use water (high $K_{max}$, low wood density) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resources, whereas species with slow traits are better protected from C losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{tlp}$, larger safety margin). Fast strategy is associated with spatially wet sites such as valleys or high rainfall [@Kursar:2009; @Oliveira:2019], but whether and how the fast-slow spectrum holds true for the moisture dynamics vertically belowground is not clear. Deep rootedness may be expected to support a fast strategy with a steady water supply unlike shallow-rooted trees. However, this may depend on the exact dynamics of water-availability by depth at a site, depending on precipitation regime and hydrology. In contrast, in pulse-driven arid systems *shallow*-rooted plants are associated with fast traits (high $K_{max}$) that maximise the use of the water-pulse, while *deep*-rooted plants show slow traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013].

@@

Growth Data: To minimise effect of light variation among trees and census intervals we selected only large trees (>= 30 cm dbh, diamater at breast height) and species whose maximum height was at least 30 m (hereafter, canopy species). We chose only those trees for which diameter (dbh) record was present across all size censuses (1990, 1995, 2000, 2005, 2010, 2015). We calculated individual tree ($i$) growth rates ($Gr_{i,t}$ (cm.yr^-1^)) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements. As tree growth rates increase with size, to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census onterval ($Resid_{i,t} = Gr_{i,t} - f(DBH_{i, t})$). The diameter model of growth was fitted based on all trees' data pooled together as there was insufficient species-specific data for large trees (SI ). We then standardised (centered and scaled) an individual tree time series ($Resid_{i,t}$) and obtained species-level (*n* = `r length(erd.sp)`) times-series of medians ($Growth_{t}$) that had data on at least three trees (median = xx). 
As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between GPP (gCm^2^day^-1^) and VPD (kPa; $GPP = f(VPD) + error$;  SI) and used standardised $f(VPD)$ (range 0-1) as $StdVPD$ in Eqs. \@ref(eq:M2)- \@ref(eq:M4).

@@

\
The model with VPD and without Leaf Fraction (Eq. \@ref(eq:M2)) had highest goodness-of fit (*R*^2^ = 0.51; *p* = 0.072), followed by followed by model without both VPD and and Leaf Fraction (Eq. \@ref(eq:M1); Fig. S\@ref(fig:other-erd-mods)). Models with leaf fraction had only three species with sufficient species-level goodness-of-fit (Eqs. \@ref(eq:M3)-\@ref(eq:M4)) versus seven for models without (Eqs. \@ref(eq:M1)-\@ref(eq:M2)). For rest of the analyses we therefore used species ERD estimates from the model with VPD and without Leaf Fraction (Eq. \@ref(eq:M2); Fig. \@ref(fig:erd-iso)).

@@

We calculated mortality rates for large trees (>= 10 cm dbh) of species with ERD estimates (all canopy species with maximum height >= 30 m) and average adbunance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`). For each of the seven census intervals ($t$) in the 34 year record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successives censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{t}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50ha plot of a species present in cen~1~ and dead in cen~2~, respectively, and t is the duration based on mean dates of cen~1~ and cen~2~. We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{i=1}^{6} {Mort_{t}}$ and, finally, species mortality anoamalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$. To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$). To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~leaf}$.

@@

\
In this forest modelled Effective Rooting Depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from 0.5 m to 13 m (Figure \@ref(fig:sp-erd)). For the subset of species for which hydraulic properties were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max,~stem}$; *R*^2^ = 0.39, *P* = 0.098, *n* = 8; Figure \@ref(fig:erd-hyd)A), less negative leaf  ($\Psi_{tlp}$); *R*^2^ = 0.6, p = 0.024, *n* = 8; Figure \@ref(fig:erd-hyd)B), greater vulnerability to embolism from cavitation with less negative ($\Psi_{88,~stem}$) (*R*^2^ = 0.72, p = 0.007, *n* = 8; Figure \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$; *R*^2^ = 0.74, p = 0.013, *n* = 8; Figure \@ref(fig:erd-hyd)D).

@@

```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus hydraulic propertiese for eight large, canopy species found on Barro Colorado Island (Panama). *R*^2^ and *P*-values are given in panel insets.[X-Axis labels to be added: $K_{max}$ units as in Fig. 2. Water Potentials are in MPa.]")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/Select_traits_depth.jpeg"))
```

@@

\
Species with deeper $ERD$ showed lower mean mortality rates across the 34 year record (1982-2015; *R*^2^ = 0.11, *P* = 0.075, *n* = `r length(erd.mrate.sp)`; Figure \@ref(fig:erd-mort)). For five out of seven census intervals, species with shallower $ERD$ had significantly higher mortality anomalies than deeper ERD species; Figure (\@ref(fig:erd-mort-anom)). Analysis of time spent under catastrophic soil drought showed that the proportion of days species $\Psi_{ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD (Figure \@ref(fig:erd-time)). Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25 year period for which soil water potential $\Psi_{soil,~z}$ was modelled (1990-2015), as soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dryseasons and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).

@@

# Discussion
\
Our analyses revealed that trees with deeper effective rooting depths (ERD) were associated higher hydraulic efficiency, but not safety: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species. Yet, deeper ERD species showed lower mortality rates than shallower ERD species over the 34 year record (1982-2015). Analysis of time spent below ($\Psi_{crit}$) in the soil zone matching ERD $\Psi_{soil,~z = ERD}$ showed that this was likely due to decreasing realised risk of hydraulic failure with depth. We thus show, for the first time, that interspecific variation in trees' effective rooting depths aligns species on a 'fast-slow' water economic spectrum by demonstrating a link between trees' realised drought experience, hydraulic architecture and long-term mortality outcomes. On the 'fast' end of the spectrum were deep ERD species with hydraulically efficient, but risky suit of traits optimised for an environment of reliable and abundant source of water at depth that actualised to a lower mortality-risk. At the slow end of the spectrum were shallow ERD species with inefficient, but safer xylem, likely optimised for high variability in $\Psi_{soil}$ in the shallow soil layers, resulting in higher variance in mortality rates across censuses as $\Psi_{soil,~z = ERD}$ crossed ERD $\Psi_{crit}$. Mortality rates of shallow ERD species may have been exacerbated by increased drought events during the study period. Previous work shows that half of the extreme dry seasons, estimated as < 1 SD annual water deficit, during 1931-2015 (85 yrs) occurred during 1982-2005 (24 yrs) of the study period (1982-2015). [@Condit.2017]

@@

Our study brings attention to the need for assessing drought-sensitivity in terms of species realised drought risk. and the usefuless of looking at species ecological strategies for that. As those are developed in response to plants historical environemnt. Plant traits hold clue for the historical environment that shaped their ecological strategy [@Reich:2014].

Our result of lower mortality in deep rooted trees, through a study period of greater recurrence of drought, is consistent with other studies, where among the three or four species studied through a centennial (Texas) or extreme (Italy) drought, deep rooted species showed lower mortality than shallow rooted species. Our result contrasts with the finding of @Chitra-Tarak:2018 who found that species identified as effectively deep rooted by their inverse model died *more* in a prolonged, centennial drought. Their hydrological model found that in the multi-annual drought deep water was exhausted for over two years, whereas our hydrological modeling at BCI finds that despite higher frequncy of droughts during our study period $\Psi_{soil}$ below the depth of 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dryseasons and drought periods, that is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa). The difference in our results are likely attributable to site differences in the precipitation regime. In their dry forest site (Mudumalai National Park, India) precipitation is low (1095 $\pm$ 296 mm; mean $\pm$ SD) compared to BCI (2687 $\pm$ 515 mm), and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (PET = 9xx $\pm$ xx mm), and thus at BCI deep soil layers are effectively recharged annually [@Chitra-Tarak:2018 and this study]. We note that the variation in mortality explained by in our study is low (*R*^2^ = 0.11, *P* = 0.075), and other causes may explain observed mortality. For example. based on a year-long study, lightening explained ~4.5% of mortality for trees >=10 cm dbh and ~40% for trees >=60 cm dbh [@Yanoviak:2019].

### Effective rooting depths at the tree community level
\
By estimating ERD for large trees of length(erd.sp) canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species species-rich tropical forests. Our chosen ERD model is an advancement over the only other inverse model of ERD (or water uptake depth index, @Chitra-Tarak:2018, *n* = 12 species) in terms of physiologically meaningful representation, fewer parameters, corroboration with tree and soil data on $\delta^2H$. Model by @Chitra-Tarak:2018 estimated six species-specific phenomenological parameters: three as root fraction for corresponding lumped vadose zone layers, and three as shape parameters for water uptake vs availability curves. For each species, they calculated daily water uptake successively from three layers--the supply, to satisfy the demand of the daily maximum evapo-transpiration, a function of PET and LAI. They regressed total supply to total demand ratio over census intervals (termed as growth coefficient) against growth (5 data points), and chose values for six parameters that best-fit (highest R^2^, that is, 2 additional parameters). Our inverse ERD model fits only two parameteres by way of regressing depth-specific model against 5 growth data points. Our model was constrained on VPD and species leaf vulnerability (PLC) curves, variables that are shown to be physiologically meaningful [add references]. We benefitted from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (SLA~lamina~ and Wood Specific Gravity) and parameters of PLC curves for which we collected data on a few species (*n* = 14), a 'hard' trait. This allowed us to scale-up PLC curves at the community scale (*n* = 284 species). Whether such scaling relationships between hard and soft traits hold true for other forests and their wider applicability remains to be seen as data on PLC curves in tropical forests are limited, but counter examples exist [see xx, yy, zz (Lawren, could you suggest references?]. We also note the availability of data on $\delta^2^H$ on trees and water sources belowground--soil/substrate matrix water in the vadoze zone or free water below the water table--as important for model calibration/validation. Model by Chitra-Tarak et al @Chitra-Tarak:2018 were not corroborated with such data as existing methods were inadequate to extract water bound to the soil and saprolite found in their dry forest study site (Jean Riotte, pers. comm.). Our model of belowground water-availability (ELM-FATES) was also advanced compared to that used by Chitra-Tarak et al [2018] and was calibrated with data on soil moisture dynamics by depth (15, 40, 100 cm), as well as, evapotranspiration and stream discharge, unlike theirs, which was calibrated on stream discharge and water table levels. We lacked data on water table levels which would have been useful to constrain water availability dynamics in the unobserved soil layers.

\
##### Future Directions for Model Development
\
Our model estimated ERD as a single value, unlike the dynamic ERD modelled by @Chitra-Tarak:2018, but future model developments should investigate seasonal dynamics of ERD, calibrated on corresponding calibration data on isotopic water sourcing depths. Our inverse ERD model was parametarised on 5-yearly growth data, with five data points over a 25 year period, which provided large degree of freedom to fit the model. Future studies could better constrain the model with higher resolution growth data such as those from dendrometers. This is especially important for modeling seasonally dynamic ERD. However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also $\pm$ changes due to hydrostatic stem flexing, whose importance increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2015]. Use of dendrometer data would thus entail accounting for stem hydrostatic flexing [@Zweifel:2018]. Two of our exploratory ERD models included seasonality in leaf area fraction, but we did not select them as they worsened the fit with growth data and evaluation data ($\delta^2H$, Fig. S\@ref(fig:other-erd-mods)). We interpret this result as that VPD and PLC curves may be adequate to explaine growth, but also acknowledge that our estimate of seasonality of leaf fraction is a tentative estimate based on leaf fall data and assumption of leaf presence during the wet season and omits inter-annual variation. Future studies could make use of data from new technologies such as drone based monitoring of species-specific leaf area fraction in species-rich forests [@Park.2019]. In forests such as BCI where understorey tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes, especially from a few growth data points, may need accounting for variation in trees' light environment over time and among trees' of the same size class, an issue we minimised by limiting ERD estimates to canopy trees. To estimate topographic variation in ERD within or across species future studies may use soil moisture dynamics from a distributed hydrological model [e.g. @Schwantes:2018]. We note that data availability on soil water retention curve ($\Psi_{soil}$ vs. VWC), hydraulic conductivity ($K_{soil}$) by depth, soil moisture by depth, stream discharge and evapotranspiration were important for effective calibration of our 1-D hydrological model, and recommend wide-spread data collection of these variables as well as water-table levels in forest-inventory sites to allow for estimation of trees' water environments. Finally, developing an inverse ERD model that accounts for movement of water within the soil-plant-atmosphere-continuum (SPAC) would be a step change from the current statistical models of ERD [this study; @Chitra-Tarak:2018]. Although data needs for parametarising such models [Christoffersen.2016tbh;@Sperry.1998] could be greater (e.g. retention curves for roots), limiting immediate community-wide application, they promise to improve our understanding of plant physiology [e.g. see @Johnson:2018fad], but which may be potentially harnessed for a community level scale-up.

>>>>

\
**Corresponding author**: 
Rutuja Chitra-Tarak
\
E: arutuj@gmail.com
\
T: +1(505) 665 1647

@@

\
Trade-offs shape how trees' make a living in their environment (ecological strategies) and also provide Dynamic Vegetation Models (DVMs) the essential constraints to represent species-rich communities under changing environment [@Koven:2020]. This entails indentifying trade-offs among environment, traits and demography. @Reich:2014 concenptualised a universal 'fast-slow' (aka productive-persistent, or acquisitive-conservative) plant economics spectrum coordinated for all resources and organs, both above- and below-ground. specifically for water observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified a fast-slow spectrum, although examples of all trade-offs observed simultaneously are rare, and belowground trait or environment linkages are weak. 'Fast' species at the productive end of the spectrum have traits with high capacity to acquire, move, store and use water (high $K_{max}$, low wood density) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resources, whereas species with slow traits are better protected from C losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{tlp}$, larger safety margin). Fast strategy is associated with spatially wet sites such as valleys or high rainfall [@Kursar:2009; @Oliveira:2019], but whether and how the fast-slow spectrum holds true for the moisture dynamics vertically belowground is not clear. Deep rootedness may be expected to support a fast strategy with a steady water supply unlike shallow-rooted trees. However, this may depend on the exact dynamics of water-availability by depth at a site, depending on precipitation regime and hydrology. In contrast, in pulse-driven arid systems *shallow*-rooted plants are associated with fast traits (high $K_{max}$) that maximise the use of the water-pulse, while *deep*-rooted plants show slow traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013].

@@

Growth Data: To minimise effect of light variation among trees and census intervals we selected only large trees (>= 30 cm dbh, diameter at breast height) and species whose maximum height was at least 30 m (hereafter, canopy species). We chose only those trees for which diameter (dbh) record was present across all size censuses (1990, 1995, 2000, 2005, 2010, 2015). We calculated individual tree ($i$) growth rates ($Gr_{i,t}$ (cm.yr^-1^)) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements. As tree growth rates increase with size, to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($Resid_{i,t} = Gr_{i,t} - f(DBH_{i, t})$). The diameter model of growth was fitted based on all trees' data pooled together as there was insufficient species-specific data for large trees (SI ). We then standardized (centered and scaled) an individual tree time series ($Resid_{i,t}$) and obtained species-level (*n* = `r length(erd.sp)`) times-series of medians ($Growth_{t}$) that had data on at least three trees (median = xx). 
As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between GPP (gCm^2^day^-1^) and VPD (kPa; $GPP = f(VPD) + error$;  SI) and used standardised $f(VPD)$ (range 0-1) as $StdVPD$ in Eqs. \@ref(eq:M2)- \@ref(eq:M4).

@@

\
The model with VPD and without Leaf Fraction (Eq. \@ref(eq:M2)) had highest goodness-of fit (*R*^2^ = 0.51; *p* = 0.072), followed by followed by model without both VPD and abundance Leaf Fraction (Eq. \@ref(eq:M1); Fig. S\@ref(fig:other-erd-mods)). Models with leaf fraction had only three species with sufficient species-level goodness-of-fit (Eqs. \@ref(eq:M3)-\@ref(eq:M4)) versus seven for models without (Eqs. \@ref(eq:M1)-\@ref(eq:M2)). For rest of the analyses we therefore used species ERD estimates from the model with VPD and without Leaf Fraction (Eq. \@ref(eq:M2); Fig. \@ref(fig:erd-iso)).

@@

We calculated mortality rates for large trees (>= 10 cm dbh) of species with ERD estimates (all canopy species with maximum height >= 30 m) and average adbunance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`). For each of the seven census intervals ($t$) in the 34-yr record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successive censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{t}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50ha plot of a species present in cen~1~ and dead in cen~2~, respectively, and t is the duration based on mean dates of cen~1~ and cen~2~. We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{i=1}^{6} {Mort_{t}}$ and, finally, species mortality anoamalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$. To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$). To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~leaf}$.

@@

\
In this forest modelled Effective Rooting Depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from 0.5 m to 13 m (Fig. \@ref(fig:sp-erd)). For the subset of species for which hydraulic properties were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max,~stem}$; *R*^2^ = 0.39, *P* = 0.098, *n* = 8; Fig. \@ref(fig:erd-hyd)A), less negative leaf  ($\Psi_{tlp}$); *R*^2^ = 0.6, p = 0.024, *n* = 8; Fig. \@ref(fig:erd-hyd)B), greater vulnerability to embolism from cavitation with less negative ($\Psi_{88,~stem}$) (*R*^2^ = 0.72, p = 0.007, *n* = 8; Fig. \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$; *R*^2^ = 0.74, p = 0.013, *n* = 8; Fig. \@ref(fig:erd-hyd)D).

@@

```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus hydraulic properties for eight large, canopy species found on Barro Colorado Island (Panama). *R*^2^ and *P*-values are given in panel insets.[X-Axis labels to be added: $K_{max}$ units as in Fig. 2. Water Potentials are in MPa.]")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/Select_traits_depth.jpeg"))
```

@@

\
Species with deeper $ERD$ showed lower mean mortality rates across the 34-yr record (1982-2015; *R*^2^ = 0.11, *P* = 0.075, *n* = `r length(erd.mrate.sp)`; Fig. \@ref(fig:erd-mort)). For five out of seven census intervals, species with shallower $ERD$ had significantly higher mortality anomalies than deeper ERD species; Fig. (\@ref(fig:erd-mort-anom)). Analysis of time spent under catastrophic soil drought showed that the proportion of days species $\Psi_{ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD (Fig. \@ref(fig:erd-time)). Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25-yrperiod for which soil water potential $\Psi_{soil,~z}$ was modelled (1990-2015), as soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).

@@

# Discussion
\
Our analyses revealed that co-occurring large canopy tree species (*n* = ` r length(erd.sp)`) with deeper effective rooting depths (ERD) were associated with higher hydraulic efficiency, but not safety: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species. Yet, deeper ERD species showed lower mortality rates than shallower ERD species over the 34-yr record (1982-2015). Analysis of time spent below ($\Psi_{crit}$) in the soil zone matching ERD $\Psi_{soil,~z = ERD}$ showed that this was likely due to decreasing realised risk of hydraulic failure with depth. We thus show, for the first time, that interspecific variation in trees' effective rooting depths aligns species on a 'fast-slow' water economic spectrum by demonstrating a link between trees' realised drought experience, hydraulic architecture and long-term mortality outcomes. On the 'fast' end of the spectrum were deep ERD species with hydraulically efficient, but risky suit of traits optimised for an environment of reliable and abundant source of water at depth that actualised to a lower mortality-risk. At the slow end of the spectrum were shallow ERD species with inefficient, but safer xylem, likely optimised for high variability in $\Psi_{soil}$ in the shallow soil layers, resulting in higher variance in mortality rates across censuses as $\Psi_{soil,~z = ERD}$ crossed ERD $\Psi_{crit}$. Mortality rates of shallow ERD species may have been exacerbated by increased drought events during the study period. Previous work shows that half of the extreme dry seasons, estimated as < 1 SD annual water deficit, during 1931-2015 (85-yr) occurred during 1982-2005 (24-yr) of the 34-yr long study period (1982-2015; @Condit.2017)

@@

Our study brings attention to the need for assessing drought-sensitivity in terms of species realised drought risk. We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins traits, that are identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through a study period of greater recurrence of drought. These results are consistent with the rare studies that studied three or four species each with known rooting or water sourcing depths and assesed their hydraulic traits and mortality rates during a centennial (Texas) or extreme (Italy) drought [@Johnson:2018fad; @Nardini:2015]. Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak:2018, who found that species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, centennial drought. Their hydrological model found that in the multi-annual drought deep water was exhausted for over two years, whereas our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period $\Psi_{soil}$ below the depth of 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dryseasons and drought periods, that is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa). The difference in our results are likely attributable to site differences in the precipitation regime. In their dry forest site (Mudumalai National Park, India) precipitation is low (1095 $\pm$ 296 mm; mean $\pm$ SD) compared to BCI (2687 $\pm$ 515 mm), and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (PET = 9xx $\pm$ xx mm), and thus at BCI deep soil layers are effectively recharged annually [@Chitra-Tarak:2018 and this study]. We note that the variation in mortality explained by in our study is low (*R*^2^ = 0.11, *P* = 0.075), and other causes may explain observed mortality. For example. based on a year-long study, lightening explained ~4.5% of mortality for trees >=10 cm dbh and ~40% for trees >=60 cm dbh [@Yanoviak:2019].

and the usefulness of looking at species ecological strategies for that. As those are developed in response to plants historical environment. Plant traits hold clue for the historical environment that shaped their ecological strategy [@Reich:2014].

### Effective rooting depths at the tree community level
\
By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species species-rich tropical forests. Our chosen ERD model is an advancement over the only other inverse model of ERD (or water uptake depth index, @Chitra-Tarak:2018, *n* = 12 species) in terms of physiologically meaningful representation, fewer parameters, corroboration with tree and soil data on $\delta^2H$. Model by @Chitra-Tarak:2018 estimated six species-specific phenomenological parameters: three as root fraction for corresponding lumped vadose zone layers, and three as shape parameters for water uptake vs availability curves. For each species, they calculated daily water uptake successively from three layers--the supply, to satisfy the demand of the daily maximum evapo-transpiration, a function of PET and LAI. They regressed total supply to total demand ratio over census intervals (termed as growth coefficient) against growth (5 data points), and chose values for six parameters that best-fit (highest R^2^, that is, 2 additional parameters). Our inverse ERD model fits only two parameters by way of regressing depth-specific model against 5 growth data points. Our model was constrained on VPD and species leaf vulnerability (PLC) curves, variables that are shown to be physiologically meaningful [add references]. We benefitted from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (SLA~lamina~ and Wood Specific Gravity) and parameters of PLC curves for which we collected data on a few species (*n* = 14), a 'hard' trait. This allowed us to scale-up PLC curves at the community scale (*n* = 284 species). Whether such scaling relationships between hard and soft traits hold true for other forests and their wider applicability remains to be seen as data on PLC curves in tropical forests are limited, but counter examples exist [see xx, yy, zz (Lawren, could you suggest references?]. We also note the availability of data on $\delta^2^H$ on trees and water sources belowground--soil/substrate matrix water in the vadoze zone or free water below the water table--as important for model calibration/validation. Model by Chitra-Tarak et al @Chitra-Tarak:2018 were not corroborated with such data as existing methods were inadequate to extract water bound to the soil and saprolite found in their dry forest study site (Jean Riotte, pers. comm.). Our model of belowground water-availability (ELM-FATES) was also advanced compared to that used by Chitra-Tarak et al [2018] and was calibrated with data on soil moisture dynamics by depth (15, 40, 100 cm), as well as, evapotranspiration and stream discharge, unlike theirs, which was calibrated on stream discharge and water table levels. We lacked data on water table levels which would have been useful to constrain water availability dynamics in the unobserved soil layers.

\
##### Future Directions for Model Development
\
Our model estimated ERD as a single value, unlike the dynamic ERD modelled by @Chitra-Tarak:2018, but future model developments should investigate seasonal dynamics of ERD, calibrated on corresponding calibration data on isotopic water sourcing depths. Our inverse ERD model was parametarised on 5-yearly growth data, with five data points over a 25-yrperiod, which provided large degree of freedom to fit the model. Future studies could better constrain the model with higher resolution growth data such as those from dendrometers. This is especially important for modeling seasonally dynamic ERD. However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also $\pm$ changes due to hydrostatic stem flexing, whose importance increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2015]. Use of dendrometer data would thus entail accounting for stem hydrostatic flexing [@Zweifel:2018]. Two of our exploratory ERD models included seasonality in leaf area fraction, but we did not select them as they worsened the fit with growth data and evaluation data ($\delta^2H$, Fig. S\@ref(fig:other-erd-mods)). We interpret this result as that VPD and PLC curves may be adequate to explaine growth, but also acknowledge that our estimate of seasonality of leaf fraction is a tentative estimate based on leaf fall data and assumption of leaf presence during the wet season and omits inter-annual variation. Future studies could make use of data from new technologies such as drone based monitoring of species-specific leaf area fraction in species-rich forests [@Park.2019]. In forests such as BCI where understory tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes, especially from a few growth data points, may need accounting for variation in trees' light environment over time and among trees' of the same size class, an issue we minimised by limiting ERD estimates to canopy trees. To estimate topographic variation in ERD within or across species future studies may use soil moisture dynamics from a distributed hydrological model [e.g. @Schwantes:2018]. We note that data availability on soil water retention curve ($\Psi_{soil}$ vs. VWC), hydraulic conductivity ($K_{soil}$) by depth, soil moisture by depth, stream discharge and evapotranspiration were important for effective calibration of our 1-D hydrological model, and recommend wide-spread data collection of these variables as well as water-table levels in forest-inventory sites to allow for estimation of trees' water environments. Finally, developing an inverse ERD model that accounts for movement of water within the soil-plant-atmosphere-continuum (SPAC) would be a step change from the current statistical models of ERD [this study; @Chitra-Tarak:2018]. Although data needs for parametarizing such models [Christoffersen.2016tbh;@Sperry.1998] could be greater (e.g. retention curves for roots), limiting immediate community-wide application, they promise to improve our understanding of plant physiology [e.g. see @Johnson:2018fad], but which may be potentially harnessed for a community level scale-up.
